Type-Safe Modular Parsing

### 2017-01-14 paper/sections/Conclusion.tex

Huang: to write a couple of sentences for a summary.

\noindent Haoyuan: to discuss future work. (some boilerplate can be avoided via metaprogramming.)

>>>>

This paper presents a solution for semantically modular parsing. It not only
enables parsers to evolve with the syntax together, but also allows them to be modularly type-checked and separately compiled.

Based on parser combinators, we identify the the algorithmic challenges of building modular parsers, and show that Packrat parsing is suitable as the underlying parsing technique. Our solution does not require advanced language features. We show that with standard OO techniques including inheritance and overriding, it is practicable to build modular parsers for OO ASTs. However, the extensibility issue of traditional OO ASTs motivates us to adopt Object Algebras for full extensibility and more useful features.

Abstracting language features as reuseable components can be achieved based on our solution, which is useful for rapid language prototyping. Our solution does not rely on particular techniques. On the contrary, it is a general framework customizable by the user. The TAPL case study shows that we can obtain considerable code reuse (69\%) than a non-modular implementation, using our modular parsing approach and language feature abstraction.

There are certainly some aspects can be improved. We observed that the glue code of composition appears to be boilerplate. Such an issue refers to family polymorphism, and we could solve it by language features or meta-programming techniques. Moreover, we can possibly adopt the Shy framework~\cite{Zhang2015} and algebra composition patterns~\cite{oliveira2013feature}, to improve the usage of Object Algebras.

For future work, we will experiment more on how open recursion contributes to extensible parsing in functional languages, by making use of laziness. It is also interesting to see that parsing, to some extent, can be viewed as a special example of \textit{unfolds}. So it is worthwhile considering to generalize our approach under certain circumstances.

### 2017-01-14 paper/sections/Inheritance.tex

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO style. On the hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{parr2013definitive}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

It is actually quite simple to let \inlinecode{pExpr} cover the newly extended case without modifying existing code. Method overriding is a standard feature which often comes with inheritance, and it allows us to explain inherited method again, such as \inlinecode{pExpr}. We can build the new parser which correctly parses \inlinecode{"1 + x"} as below.

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple inheritance.
Scala has a linearized-style multiple inheritance for traits~\cite{odersky2004overview}. This can be very helpful when composing several languages, which
may have been independently developed. Suppose now
we want to compose the parsers for expressions from pre-defined
languages \inlinecode{LanguageA} and \inlinecode{LanguageB} using
alternative. The new parser can be built by inheriting both parsers
at the same time:

>>>>

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO style~\cite{wadler1998expression}. On the hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{parr2013definitive}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

It is actually quite simple to let \inlinecode{pExpr} cover the newly
extended case without modifying existing code. Method overriding is a
standard feature which often comes with inheritance, and it allows us
to redefined an inherited method, such as \inlinecode{pExpr}. We can
build the new parser which correctly parses \inlinecode{"1 + x"} 
through overriding:

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple
inheritance.  Scala has a linearized-style multiple inheritance for
traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. 
Suppose now we want to compose the parsers for expressions
from pre-defined languages \inlinecode{LanguageA} and
\inlinecode{LanguageB} using alternative. The new parser can be built
by inheriting both parsers at the same time:
